---
title: "R Notebook"
output: html_notebook
---
```{r}
#! /usr/bin/env Rscript

# Author(s): Ignatius Pang
# Email: ipang@cmri.org.au
# Children’s Medical Research Institute, finding cures for childhood genetic diseases
```

```{r}

options(knitr.duplicate.label = "allow")


#Test if BioManager is installed 
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
   BiocManager::install(version = "3.12")
}

# load pacman package manager
if(!require(pacman)){
    install.packages("pacman")
    library(pacman)
}
p_load(optparse)
p_load(tictoc)
```
```{r}
tic()
```


```{r}
p_load(tidyverse)
p_load(vroom)
p_load(magrittr)
p_load(rlang)
p_load( janitor)
p_load(UniProt.ws)
p_load(biomaRt)
p_load(GO.db)
p_load(ProteomeRiver)
```


```{r}
list_of_sp_columns <- c("EXISTENCE", "SCORE", "REVIEWED", "GENENAME", "PROTEIN-NAMES", "LENGTH", "ENSEMBL", "GO-ID", "KEYWORDS")
```

```{r}

group_pattern <- "NR"
## Directories management 
base_dir <- here::here()
data_dir <- file.path( base_dir, "Data", "Abundance_Data", "P90")
results_dir <- file.path(base_dir, "Results",  paste0(group_pattern, "90"),  "Proteins", "DE_Analysis")
source_dir <- file.path(base_dir, "Source")

```

## Command to convert the R markdown file to a R file for use in command line. 
knitr::purl( file.path(  source_dir, "Shared",  "annot_proteins_cmd.Rmd"), 
             output=file.path( source_dir,  "Shared",   "annot_proteins_cmd.R") ) 
       

```{r}

input_wide_file <- file.path(results_dir, "de_proteins_wide.tsv" ) 
input_long_file <- file.path(results_dir, "de_proteins_long.tsv" ) 
output_long_file <- file.path(results_dir, "de_proteins_longer_annot.tsv" ) 
output_wide_file <- file.path(results_dir, "de_proteins_wider_annot.tsv" ) 
ids_file <- file.path( results_dir, "cleaned_accession_to_protein_group.tab")
taxonomy_id <- 10090
raw_counts_file <- file.path(data_dir, "ALPK1-set1proteinGroups.txt")


command_line_options <- commandArgs(trailingOnly = TRUE)
if( length(command_line_options ) > 0 ) {
  parser <- OptionParser(add_help_option =TRUE)

  parser <- add_option(parser, c( "--tax-id"), type="integer", default="", dest = "taxonomy_id",
                       help="The NCBI taxonomy ID of the organism being investigated (e.g. M. musculus=10090, H. sapien=9606).",
                       metavar="integer")     
  
  parser <- add_option(parser, c( "--output-dir"), type="character", default="", dest = "results_dir",
                       help="Directory path for all results files.",
                       metavar="string")   
  
  parser <- add_option(parser, c("--input-wide"), type="character", default="", dest = "input_wide_file",
                       help="Results table with values in wider format.",
                       metavar="string")   
  
  parser <- add_option(parser, c("--input-long"), type="character", default="", dest = "input_long_file",
                       help="Results table with values in longer format.",
                       metavar="string") 
  
  parser <- add_option(parser, c("--ids"), type="character", default="", dest = "ids_file",
                       help="File to link the cleaned list of accessions to the original list of protein groups from MaxQuant file. Also, contains the MaxQuant output row ID.",
                       metavar="string") 
  
  parser <- add_option(parser, c("--raw-counts"), type="character", default="", dest = "raw_counts_file",
                       help="Input file with the protein abundance data.",
                       metavar="string")   
  
  parser <- add_option(parser, c( "--output-wide"), type="character", default="", dest = "output_wide_file",
                       help="Results table with values in wider format.",
                       metavar="string")   
  
  parser <- add_option(parser, c( "--output-long"), type="character", default="", dest = "output_long_file",
                       help="Results table with values in longer format.",
                       metavar="string")   


  print(command_line_options)
  
  cmd_arguments <- parse_args(parser)
  
  print(cmd_arguments)  

  input_wide_file <- cmd_arguments$input_wide_file
  input_long_file <- cmd_arguments$input_long_file
  output_wide_file <- cmd_arguments$output_wide_file
  output_long_file <- cmd_arguments$output_long_file
  results_dir <- cmd_arguments$results_dir
  taxonomy_id <- cmd_arguments$taxonomy_id
  ids_file  <- cmd_arguments$ids_file
  raw_counts_file <- cmd_arguments$raw_counts_file

  if( results_dir == "" ) { stop("No value for --output-dir was supplied.") }
  if( group_pattern == "" ) { stop("No value for --group-pattern was supplied.") }
  if( input_wide_file == "" ) { stop("No value for --input-wide was supplied. ") }
  if( input_long_file == "" ) { stop("No value for --input-long was supplied. ") }
  if( output_wide_file == "" ) { stop("No value for --output-wide was supplied. ") }
  if( output_long_file == "" ) { stop("No value for --output-long was supplied. ") }
  if( taxonomy_id == "" ) { stop("No value for --tax_id was supplied. ") }
  if( ids_file == "" ) { stop("No value for --ids was supplied.") }
  if( raw_counts_file == "" ) { stop("No value for -raw-counts was supplied.") }
  
}

create_dir_if_not_exists(results_dir)
de_proteins_wider <- vroom::vroom( input_wide_file ) 
de_proteins_longer <- vroom::vroom( input_long_file ) 
ids_tbl <- vroom::vroom( ids_file ) 
dat_tbl <- vroom::vroom( raw_counts_file )
dat_cln <-  janitor::clean_names(dat_tbl) %>%
  dplyr::rename( gene_names_maxquant = "gene_names")
colnames(dat_cln) <- str_replace(colnames(dat_cln), "_i_ds", "_ids" )

```
## Download Reactome annotation file 
```{r}

print("Download Reactome UniProt to pathways file.")

temp_file <- tempfile()

reactome_file <- download.file(url="https://reactome.org/download/current/UniProt2Reactome.txt", destfile=temp_file)

reactome_map <- vroom::vroom( temp_file , 
                              col_names = c("uniprot_acc", "reactome_id", "url", "reactome_term", "evidence", "organism") )




```



```{r}

print("Get the best UniProt accession per row.")

uniprot_acc_tbl <- de_proteins_wider %>%
  mutate( uniprot_acc_copy = uniprot_acc ) %>%
  separate_rows(uniprot_acc_copy, sep=":" ) %>%
  mutate( join_uniprot_acc = clean_isoform_number(uniprot_acc_copy)) %>%
  dplyr::distinct( uniprot_acc, join_uniprot_acc) %>%
  group_by( uniprot_acc) %>%
  mutate( acc_order_id = row_number()) %>% 
  ungroup


```


```{r}
print("Download information from UniProt.")
uniprot_dat <- NA
up <- NA
if( ! file.exists( file.path(results_dir, "uniprot_data.RDS"))) {

  if( is.na(up)) {
    up <- UniProt.ws(taxId=taxonomy_id )
  } 

 # keytypes(up)
  
  uniprot_dat <- batch_query_evidence(uniprot_acc_tbl, join_uniprot_acc, uniprot_handle=up, 
                                      uniprot_columns = list_of_sp_columns)
  
  saveRDS( uniprot_dat, file.path(results_dir, "uniprot_data.RDS"))
  
} else {
  uniprot_dat <- readRDS( file.path(results_dir, "uniprot_data.RDS"))
}

```


```{r eval=FALSE}
 # uniprot_dat <- batch_query_evidence(uniprot_acc_tbl %>% head(100), best_uniprot_acc, uniprot_handle=up, 
 #                                      uniprot_columns = list_of_sp_columns)
```

```{r}

print("Merge with Gene Ontology terms.")
goterms <- Term(GOTERM)
gotypes <- Ontology(GOTERM)


tic()
uniprot_dat_cln <- uniprot_go_id_to_term( uniprot_dat, sep="; ", goterms, gotypes  )
toc()



uniprot_dat_multiple_acc <- uniprot_acc_tbl %>%
  left_join( uniprot_dat_cln, by=c("join_uniprot_acc" = "UNIPROTKB") )   %>%
  arrange( uniprot_acc, acc_order_id) %>%
  group_by(uniprot_acc ) %>%
  summarise( across( .cols=setdiff( colnames( uniprot_dat_cln), "UNIPROTKB")   , ~paste(., collapse=":"))   ) %>%
  ungroup()   %>%
  dplyr::rename( UNIPROT_GENENAME = "GENENAME")


uniprot_dat_multiple_acc



```



```{r}
print("Add reactome pathways annotation.")
reactome_term_tbl <- uniprot_acc_tbl %>%
  left_join( reactome_map, by=c("join_uniprot_acc" = "uniprot_acc") )   %>%
  dplyr::filter(reactome_term != "NA" ) %>%
  group_by(uniprot_acc, join_uniprot_acc) %>%
  summarise( reactome_term = paste(reactome_term, collapse="; ") ) %>%
  ungroup()     %>%
  mutate(reactome_term = str_replace_all( reactome_term , ":", "-")) %>%
  group_by(uniprot_acc ) %>%
  summarise( reactome_term = paste(reactome_term, collapse=":") ) %>%
  ungroup()   


```


```{r}
print("Output wider format results table with protein annotation.")
de_proteins_wider_annot <- de_proteins_wider %>%
  left_join( ids_tbl, by=c("uniprot_acc" = "uniprot_acc") ) %>% 
  left_join( uniprot_dat_multiple_acc, by = c("uniprot_acc" = "uniprot_acc") ) %>%
  left_join( reactome_term_tbl, by = c("uniprot_acc" = "uniprot_acc"))  %>%
  left_join( dat_cln, by=c("maxquant_row_id" = "id",
                           "protein_ids" = "protein_ids"))

head( de_proteins_wider_annot ) 

vroom::vroom_write(de_proteins_wider_annot, path=output_wide_file ) 

```


```{r}
print("Output longer format results table with protein annotation.")
de_proteins_longer_annot <- de_proteins_longer %>%
  left_join( ids_tbl, by=c("uniprot_acc" = "uniprot_acc") ) %>% 
  left_join( uniprot_dat_multiple_acc, by = c("uniprot_acc" = "uniprot_acc") ) %>%
  left_join( reactome_term_tbl, by = c("uniprot_acc" = "uniprot_acc"))  %>%
  left_join( dat_cln, by=c("maxquant_row_id" = "id",
                           "protein_ids" = "protein_ids"))

head( de_proteins_longer_annot )

vroom::vroom_write(de_proteins_longer_annot, path=output_long_file ) 

```

```{r}
toc()
sessionInfo()
```


## Available uniprot keys
[1] "AARHUS/GHENT-2DPAGE"        "AGD"                        "ALLERGOME"                  "ARACHNOSERVER"              "BIOCYC"                     "CGD"                       
 [7] "CLEANEX"                    "CONOSERVER"                 "CYGD"                       "DICTYBASE"                  "DIP"                        "DISPROT"                   
[13] "DMDM"                       "DNASU"                      "DRUGBANK"                   "ECHOBASE"                   "ECO2DBASE"                  "ECOGENE"                   
[19] "EGGNOG"                     "EMBL/GENBANK/DDBJ"          "EMBL/GENBANK/DDBJ_CDS"      "ENSEMBL"                    "ENSEMBL_GENOMES"            "ENSEMBL_GENOMES PROTEIN"   
[25] "ENSEMBL_GENOMES TRANSCRIPT" "ENSEMBL_PROTEIN"            "ENSEMBL_TRANSCRIPT"         "ENTREZ_GENE"                "EUHCVDB"                    "EUPATHDB"                  
[31] "FLYBASE"                    "GENECARDS"                  "GENEFARM"                   "GENEID"                     "GENENAME"                   "GENETREE"                  
[37] "GENOLIST"                   "GENOMEREVIEWS"              "GENOMERNAI"                 "GERMONLINE"                 "GI_NUMBER*"                 "H-INVDB"                   
[43] "HGNC"                       "HOGENOM"                    "HPA"                        "HSSP"                       "KEGG"                       "KO"                        
[49] "LEGIOLIST"                  "LEPROMA"                    "MAIZEGDB"                   "MEROPS"                     "MGI"                        "MIM"                       
[55] "MINT"                       "NEXTBIO"                    "NEXTPROT"                   "OMA"                        "ORPHANET"                   "ORTHODB"                   
[61] "PATRIC"                     "PDB"                        "PEROXIBASE"                 "PHARMGKB"                   "PHOSSITE"                   "PIR"                       
[67] "POMBASE"                    "PPTASEDB"                   "PROTCLUSTDB"                "PSEUDOCAP"                  "REACTOME"                   "REBASE"                    
[73] "REFSEQ_NUCLEOTIDE"          "REFSEQ_PROTEIN"             "RGD"                        "SGD"                        "TAIR"                       "TCDB"                      
[79] "TIGR"                       "TUBERCULIST"                "UCSC"                       "UNIPARC"                    "UNIPATHWAY"                 "UNIPROTKB"                 
[85] "UNIPROTKB_ID"               "UNIREF100"                  "UNIREF50"                   "UNIREF90"                   "VECTORBASE"                 "WORLD-2DPAGE"              
[91] "WORMBASE"                   "WORMBASE_PROTEIN"           "WORMBASE_TRANSCRIPT"        "XENBASE"                    "ZFIN"          


## Available uniprot data columns
  [1] "3D"                         "AARHUS/GHENT-2DPAGE"        "AGD"                        "ALLERGOME"                  "ARACHNOSERVER"              "BIOCYC"                    
  [7] "CGD"                        "CITATION"                   "CLEANEX"                    "CLUSTERS"                   "COMMENTS"                   "CONOSERVER"                
 [13] "CYGD"                       "DATABASE(PDB)"              "DATABASE(PFAM)"             "DICTYBASE"                  "DIP"                        "DISPROT"                   
 [19] "DMDM"                       "DNASU"                      "DOMAIN"                     "DOMAINS"                    "DRUGBANK"                   "EC"                        
 [25] "ECHOBASE"                   "ECO2DBASE"                  "ECOGENE"                    "EGGNOG"                     "EMBL/GENBANK/DDBJ"          "EMBL/GENBANK/DDBJ_CDS"     
 [31] "ENSEMBL"                    "ENSEMBL_GENOMES"            "ENSEMBL_GENOMES PROTEIN"    "ENSEMBL_GENOMES TRANSCRIPT" "ENSEMBL_PROTEIN"            "ENSEMBL_TRANSCRIPT"        
 [37] "ENTREZ_GENE"                "ENTRY-NAME"                 "EUHCVDB"                    "EUPATHDB"                   "EXISTENCE"                  "FAMILIES"                  
 [43] "FEATURES"                   "FLYBASE"                    "FUNCTION"                   "GENECARDS"                  "GENEFARM"                   "GENEID"                    
 [49] "GENENAME"                   "GENES"                      "GENETREE"                   "GENOLIST"                   "GENOMEREVIEWS"              "GENOMERNAI"                
 [55] "GERMONLINE"                 "GI_NUMBER*"                 "GO"                         "GO-ID"                      "H-INVDB"                    "HGNC"                      
 [61] "HOGENOM"                    "HPA"                        "HSSP"                       "ID"                         "INTERACTOR"                 "INTERPRO"                  
 [67] "KEGG"                       "KEYWORD-ID"                 "KEYWORDS"                   "KO"                         "LAST-MODIFIED"              "LEGIOLIST"                 
 [73] "LENGTH"                     "LEPROMA"                    "MAIZEGDB"                   "MEROPS"                     "MGI"                        "MIM"                       
 [79] "MINT"                       "NEXTBIO"                    "NEXTPROT"                   "OMA"                        "ORGANISM"                   "ORGANISM-ID"               
 [85] "ORPHANET"                   "ORTHODB"                    "PATHWAY"                    "PATRIC"                     "PDB"                        "PEROXIBASE"                
 [91] "PHARMGKB"                   "PHOSSITE"                   "PIR"                        "POMBASE"                    "PPTASEDB"                   "PROTCLUSTDB"               
 [97] "PROTEIN-NAMES"              "PSEUDOCAP"                  "REACTOME"                   "REBASE"                     "REFSEQ_NUCLEOTIDE"          "REFSEQ_PROTEIN"            
[103] "REVIEWED"                   "RGD"                        "SCORE"                      "SEQUENCE"                   "SGD"                        "SUBCELLULAR-LOCATIONS"     
[109] "TAIR"                       "TAXONOMIC-LINEAGE"          "TCDB"                       "TIGR"                       "TOOLS"                      "TUBERCULIST"               
[115] "UCSC"                       "UNIPARC"                    "UNIPATHWAY"                 "UNIPROTKB"                  "UNIPROTKB_ID"               "UNIREF100"                 
[121] "UNIREF50"                   "UNIREF90"                   "VECTORBASE"                 "VERSION"                    "VIRUS-HOSTS"                "WORLD-2DPAGE"              
[127] "WORMBASE"                   "WORMBASE_PROTEIN"           "WORMBASE_TRANSCRIPT"        "XENBASE"                    "ZFIN"    


# ```{r}
# # dataset="hsapiens_gene_ensembl"
# # listEnsembl()
# # mart<- useEnsembl(biomart="genes")
# # listDatasets(mart)
# # searchDatasets(mart = mart, pattern = "hsapiens")
# # searchAttributes(mart = mart, pattern = "uniprot")
# # listFilters(mart)
# # listAttributes(mart)
# 
# mart<- useEnsembl(biomart="genes", dataset="mmusculus_gene_ensembl")
# 
# 
# 	
# 	sp_annot <- getBM(filters= c( "uniprotswissprot"),
# 						attributes= c("uniprotswissprot", "ensembl_gene_id", "go_id", "kegg_enzyme", "reactome"),
# 						values= uniprot_acc_tbl %>% pull(uniprot_acc),
# 						mart= mart)
# 
# 
# 	trembl_annot <- getBM(filters= c( "uniprotsptrembl"),
# 						attributes= c("uniprotsptrembl", "ensembl_gene_id", "go_id", "kegg_enzyme", "reactome"),
# 						values= uniprot_acc_tbl %>% pull(uniprot_acc),
# 						mart= mart)
# 	
# 	
# 	sp_annot %>%
# 	  distinct(uniprotsptrembl)
# ```